Microxanox: an R package for simulating an aquatic $MICR$obial ecosystem that can occupy $OX$ic or $ANOX$ic states.

### 2022-05-10 inst/manuscript/manuscript_microxanox.Rmd
Additional references in bibliograhy and beautified graphs.

Mathematical models play a key role in the development of understanding about how ecosystems work and how they respond to environmental changes [@REF_NEEDED]
- Chaparro-Pedraza, P.C. (2021). Fast environmental change and eco-evolutionary feedbacks can drive regime shifts in ecosystems before tipping points are crossed. Proc. R. Soc. B., 288, 20211192.
- Binzer, A., Guill, C., Rall, B.C. & Brose, U. (2016). Interactive effects of warming, eutrophication and size structure: impacts on biodiversity and food-web structure. Global Change Biology, 22, 220–227.
- Vasseur, D. a & McCann, K.S. (2005). A mechanistic approach for modeling temperature-dependent consumer-resource dynamics. The American Naturalist, 166, 184–198.

. They are also critical for developing hypotheses that can then be tested in experimental studies. One area of ecology in which simple models have played a very influential role is the area of how ecosystems respond to gradual environmental change [@REF_NEEDED].
- Scheffer, M., Carpenter, S., Foley, J. a, Folke, C. & Walker, B. (2001). Catastrophic shifts in ecosystems. Nature, 413, 591–6.

It is conceivable that an ecosystem state, such as the total biomass of a particular type of bacteria, may remain unchanged when something about its environment changes, such as the temperature. It is possible that the ecosystem state changes gradually itself. It is also possible that the ecosystem state may change abruptly into a new state that is difficult to recover from [@REF_NEEDED]
- Scheffer, M., Carpenter, S., Foley, J. a, Folke, C. & Walker, B. (2001). Catastrophic shifts in ecosystems. Nature, 413, 591–6.

This possibility for abrupt, perhaps catastrophic changes that are difficult to reverse causes considerable concern [@REF_NEEDED]. 
- Collins, S.L., Nippert, J.B., Blair, J.M., Briggs, J.M., Blackmore, P. & Ratajczak, Z. (2021). Fire frequency, state change and hysteresis in tallgrass prairie. Ecology Letters, 24, 636–647.
- Northrop, A.C., Avalone, V., Ellison, A.M., Ballif, B.A. & Gotelli, N.J. (n.d.). Clockwise and counterclockwise hysteresis characterize state changes in the same aquatic ecosystem. Ecology Letters, n/a.
- Vandermeer, J. & Perfecto, I. (2019). Hysteresis and critical transitions in a coffee agroecosystem. PNAS, 116, 15074–15079.

@@

One limitation of the study by Bush et al (2017) was limited biodiversity in the ecosystem model. Specifically, there were three functional groups of bacteria, but within each of these groups, there was no biodiversity. This then leaves open the question of if and how biodiversity within functional groups, in their model, affects the ecosystem response to environmental change. This limitation is not limited to the study of Bush et al (2017): there are few if any studies of the effects of biodiversity on abrupt transitions between ecosystem states.

We decided to fill this research gap, by making a simulation study of how within functional group biodiversity affects ecosystem responses to environmental change @Limberger_et_al_2022, and to base our work on the work and model of Bush et al (2017). It was with this goal in mind that we developed the *microxanox* package. The first stage of development was to write code from scratch (as there was no available code to start from) and to confirm that this new implementation would reproduce the previously published results. The resulting reproduction is available as one of the package vignettes: [vignette Partial reproduction of Bush et al](LINK NEEDED).

@@

The *microxanox* package is for simulating a three functional group system (*CB*: cyanobacteria, *PB*: phototrophic sulfur bacteria, and *SB*: sulfate-reducing bacteria) with four chemical substrates (*P*: phosphorus, *O*: oxygen, *SR*: reduced sulfur, *SO*: oxidized sulfur). It includes feedback between biogeochemical processes and is based on [@Bush2017] (See [@Bush2017] for a detailed discussion of the model). At the core of the simulations is a set of ordinary differential equations (specified in the function `bushplus_dynamic_model()`, though users do not need to call this function directly). There are functions for running an individual simulations and for running a set of simulations across, for example, a range of environmental conditions.

@@

The framework used when writing this package aimed to maximise simplicity for the user, and to make it easy to reproduce results (see [@REF_NEEDED] for examples). As such, all the parameters needed to run a simulation or find a stable state are contained in a single object (which can easily be created using included functions). This parameter object is given to a function that runs the simulations and returns the results. The returned results object is identical to the parameter object but with an additional slot named `results`, which contains the results of the run. Thus the returned results object contains the simulation conditions (parameters) as well as the rsults, and can be used to run the simulation again. This promotes reproducibility and makes incremental changes of individual parameters and re-running the simulations straightforward.

@@

A simulation is run using the `run_simulation()` function. In this function, the ODEs are solved using the function `ode` in the package deSolve package [@Soetaert2010]. The `run_simulation()` function needs only one argument - an object as created by the function `new_runsim_parameter()`. The parameter object returned by `new_runsim_parameter()` contains among other things the `strain_parameter` object, which can be created by the function `new_strain_parameter()`. For a detailed description of the parameter objects, their meaning and how they are created and have values set and changed please see the *User Guide* which accompanies the package or is available at [User Guide](@LINK_NEEDED)

@@

Two methods for finding steady states are implemented. The first runs a separate simulation for each combination of starting conditions and oxygen diffusivity (let us term this the *Replication method*). This is the method used in the Bush et al (2017) study. The second runs only two simulations, with step-wise and slowly temporally increasing oxygen diffusivity, and the other with step-wise and slowly decreasing or decreasing oxygen diffusivity. During this temporal environmental change, the state of the system was recorded just before change to a new oxygen diffusivity (let us term this the *Temporal method*). We implemented two methods since there is no definitive best method, and in order to check if results were sensitive to choice of method.

@@

From the raw results returned can be extracted summary quantity about how the ecoystem stable states respond to environmental change. The function `get_stability_measures()` returns quantities such as the amount of environmental change required for the system to abruptly change to a different state.

@@

**Rainer... the figure needs beautifying please.**

```{r uc1, echo = FALSE,  include=TRUE, fig.cap='The temporal dynamics of the ecosystem model when an environmental condition (here parameter *a*, the oxygen diffusivity) changes. Plot of the stable states of the simulation runs under different oxygen diffusivity.', fig.align="center", out.width = 350 }
knitr::include_graphics("figures/ug_temporal_state_switching.png.png")
```

@@

The package also contains a function to extract summary features of ecosystem responses to environmental change, such as the amount of hysteresis displayed by the ecosystem. Hysteresis is a key feature of ecosystem responses to environemtal change, because it is related to how difficult it can be to reserve the effects of environmental change [@REF_NEEDED Scheffer, M., Carpenter, S., Foley, J. a, Folke, C. & Walker, B. (2001). Catastrophic shifts in ecosystems. Nature, 413, 591–6.
]. The amount of hysteresis was measured here as the extent of the environmental condition for which there were two stable states. I.e. it was the extent of the environmental conditions for which historical conditions play an important role in determining the current system state (a definition of hysteresis). The results show that the amount of hysteresis depends greatly on the combinations of organisms present (Figure \@ref(fig:uc2)). For example, with only the CB (cyanobacteria) present, there was no hysteresis. In contrast, the presence of both CB and SB (sulfate reducing bacteria) led to a large amount of hysteresis. (These results are also given in the Partial Reproduction vignette.)

**Rainer... the figure needs beautifying please.**

```{r uc2, echo = FALSE,  include=TRUE, fig.cap='The amount of hysteresis depends on the combination of types of organisms present.', fig.align="center", out.width = 350 }
knitr::include_graphics("figures/user_guide_hysteresis.png")
```

@@

As discussed in the Introduction section, the package was largely motivated by the question of how biodiversity influences ecosystem responses to environmental change. Extensive results concerning this question are given in a separate publication @REF_NEEDED. Here we describe one of the results, which is that having biodiversity in a functional group can allow state changes to occur that otherwise would not have. I.e. biodiversity can qualitatively change the state of the ecosystem.

Biodiversity was added to the functional groups using the 'new_strain_parameter()' function to create a parameter set with multiple species per functional group (albeit all with identical features) and then to add variability among the species by calling the 'add_strain_var()' function. This function takes an already existing parameter set and adds the specified about of variation. The new parameter object is then used as before.

@@

**Rainer... please beautify the figure as required. It comes from the Limberger et al supplement, Figure 5.4**

```{r uc3, echo = FALSE,  include=TRUE, fig.cap='The dynamics of the ecosystem when there are two species in each functional group, and some variation (diversity) in species parameters.', fig.align="center", out.width = 350 }
knitr::include_graphics("figures/uc3_2strain_temporal_switch.png")
```

@@

*microxanox* allows the simulation, visualisation, and analysis of a model of a microbial ecosystem while allowing variation in the amount of diversity containing in each of the functional groups of organisms present. It has been used for the research described in another paper of ours that is one of the first investigations of the effects of diversity on ecosystem resilience @REF_NEEDED. In that paper, we show that diversity can have large and important effects of ecosystem responses, highlighting the need for models such as ours, within which one can easily manipulate the amount of biodiversity. The 'microxanox' package has also been used to reproduce the results of the paper that inspired the package development @Bush2017.

The package greatly lowers the amount of work required in further investigations of the specific ecosystem modelled. There has, for example, been quite limited investigation of how biodiversity influences the short-term responses of the modelled ecosystem to environmental change. Likewise, the package could be used to power an investigation of the effects of biodiversity on the usefulness of early warning signals of abrupt ecosystem change @REF_NEEDED ->>> Scheffer, M., Bascompte, J., Brock, W.A., Brovkin, V., Carpenter, S.R., Dakos, V., et al. (2009). Early-warning signals for critical transitions. Nature, 461, 53–59.
]. In addition this package could be used as a template for the implementation for developing models of other types of ecosystems and organism. By doing so, other models can profit from the overall framework used, and the reproducibility aspects as well as the flexibility implemented.

>>>>

Mathematical models play a key role in the development of understanding about how ecosystems work and how they respond to environmental changes [@Binzer2016a; @Chaparro-Pedraza2021; @Vasseur2005]. They are also critical for developing hypotheses that can then be tested in experimental studies. One area of ecology in which simple models have played a very influential role is the area of how ecosystems respond to gradual environmental change [@REF_NEEDED].
- Scheffer, M., Carpenter, S., Foley, J. a, Folke, C. & Walker, B. (2001). Catastrophic shifts in ecosystems. Nature, 413, 591–6.

It is conceivable that an ecosystem state, such as the total biomass of a particular type of bacteria, may remain unchanged when something about its environment changes, such as the temperature. It is possible that the ecosystem state changes gradually itself. It is also possible that the ecosystem state may change abruptly into a new state that is difficult to recover from [@Scheffer2001].

This possibility for abrupt, perhaps catastrophic changes that are difficult to reverse causes considerable concern [@Collins2021; @Northrop2021; @Vandermeer2019].

@@

One limitation of the study by @Bush2017 was limited biodiversity in the ecosystem model. Specifically, there were three functional groups of bacteria, but within each of these groups, there was no biodiversity. This then leaves open the question of if and how biodiversity within functional groups, in their model, affects the ecosystem response to environmental change. This limitation is not limited to the study of @Bush2017, there are few if any studies of the effects of biodiversity on abrupt transitions between ecosystem states.

We decided to fill this research gap, by making a simulation study of how within functional group biodiversity affects ecosystem responses to environmental change @Limberger2022, and to base our work on the work and model of @Bush2017. It was with this goal in mind that we developed the *microxanox* package. The first stage of development was to write code from scratch (as there was no available code to start from) and to confirm that this new implementation would reproduce the previously published results. The resulting reproduction is available as one of the package vignettes: [vignette Partial reproduction of Bush et al](LINK NEEDED).

@@

The *microxanox* package is for simulating a three functional group system (*CB*: cyanobacteria, *PB*: phototrophic sulfur bacteria, and *SB*: sulfate-reducing bacteria) with four chemical substrates (*P*: phosphorus, *O*: oxygen, *SR*: reduced sulfur, *SO*: oxidized sulfur). It includes feedback between biogeochemical processes and is based on @Bush2017 (See @Bush2017 for a detailed discussion of the model). At the core of the simulations is a set of ordinary differential equations (specified in the function `bushplus_dynamic_model()`, though users do not need to call this function directly). There are functions for running individual simulations and for running a set of simulations across, for example, a range of environmental conditions.

@@

The framework used when writing this package aimed to maximise simplicity for the user, and to make it easy to reproduce results (see the supplement to @Limberger2022 for an example of how this is used). As such, all the parameters needed to run a simulation or find a stable state are contained in a single object (which can easily be created using included functions). This parameter object is given to a function that runs the simulations and returns the results. The returned results object is identical to the parameter object but with an additional slot named `results`, which contains the results of the run. Thus the returned results object contains the simulation conditions (parameters) as well as the rsults, and can be used to run the simulation again. This promotes reproducibility and makes incremental changes of individual parameters and re-running the simulations straightforward.

@@

A simulation is run using the `run_simulation()` function. In this function, the ODEs are solved using the function `ode()` in the package *deSolve* package [@Soetaert2010]. The `run_simulation()` function needs only one argument - an object as created by the function `new_runsim_parameter()`. The parameter object returned by `new_runsim_parameter()` contains among other things the `strain_parameter` object, which can be created by the function `new_strain_parameter()`. For a detailed description of the parameter objects, their meaning and how they are created and have values set and changed please see the *User Guide* which accompanies the package or is available at [User Guide](@LINK_NEEDED)

@@

Two methods for finding steady states are implemented. The first runs a separate simulation for each combination of starting conditions and oxygen diffusivity (let us term this the *Replication method*). This is the method used in the @Bush2017 study. The second runs only two simulations, with step-wise and slowly temporally increasing oxygen diffusivity, and the other with step-wise and slowly decreasing or decreasing oxygen diffusivity. During this temporal environmental change, the state of the system was recorded just before change to a new oxygen diffusivity (let us term this the *Temporal method*). We implemented two methods since there is no definitive best method, and in order to check if results were sensitive to choice of method.

@@

From the raw results returned, summary measures about how the ecoystem stable states respond to environmental change can be extracted. The function `get_stability_measures()` returns quantities such as the amount of environmental change required for the system to abruptly change to a different state.

@@

```{r uc1, echo = FALSE,  include=TRUE, fig.cap='The temporal dynamics of the ecosystem model when an environmental condition (here parameter *a*, the oxygen diffusivity) changes. Plot of the stable states of the simulation runs under different oxygen diffusivity.', fig.align="center", out.width = 350}
fig <- "./figures/gen_uc1_partrep_temporal_state_switching.pdf"
if (!file.exists(fig)){
  library(ggplot2)
  library(microxanox)
  x <- readRDS("./data/hlh_with_defaults.RDS")
  p <- plot_dynamics(x)
  p[[1]] <- p[[1]] + 
    xlab(NULL) + 
    ylab(expression(log[10](density)~(cells~L^{-1}))) +
    theme_bw()
  p[[2]] <- p[[2]] + 
    xlab(NULL) + 
    ylab(expression(log[10](density)~(cells~L^{-1}))) +
    theme_bw()
  p[[3]] <- p[[3]] + 
    xlab(NULL) + 
    ylab(expression(log[10](density)~(cells~L^{-1}))) +
    theme_bw()
  p[[4]]$data <- p[[4]]$data %>% dplyr::filter(species != "a")
  p[[4]]$data$species <- p[[4]]$data$species %>%
    dplyr::recode(SO = "Oxidised sulfur", O = "Oxygen", SR = "Sulfide", P = "Phosphorus")
  p[[4]] <- p[[4]] +
    xlab("time (h)") + 
    ylab(expression(log[10](concentration)~(µM))) + 
    labs(colour = "substrate") +
    theme_bw() +
    xlab(expression(time~(hours)))
  ggsave(
    filename = fig,
    plot = p,
    device = "pdf",
    width = 8,
    height = 8
  )  
}
knitr::include_graphics(fig)
```

@@

The package also contains a function to extract summary features of ecosystem responses to environmental change, such as the amount of hysteresis displayed by the ecosystem. Hysteresis is a key feature of ecosystem responses to environmental change, because it is related to how difficult it can be to reserve the effects of environmental change [@Scheffer2001]. The amount of hysteresis was measured here as the extent of the environmental condition for which there were two stable states. I.e. it was the extent of the environmental conditions for which historical conditions play an important role in determining the current system state (a definition of hysteresis). The results show that the amount of hysteresis depends greatly on the combinations of organisms present (Figure \@ref(fig:uc2)). For example, with only the CB (cyanobacteria) present, there was no hysteresis. In contrast, the presence of both CB and SB (sulfate reducing bacteria) led to a large amount of hysteresis. (These results are also given in the Partial Reproduction vignette.)

**@Rainer... the figure needs beautifying please.**
**@Owen: is this OK like this?**

```{r uc2, echo = FALSE,  include=TRUE, fig.cap='The amount of hysteresis depends on the combination of types of organisms present.', fig.align="center", out.width = 350 }
fig <- "./figures/gen_uc2_user_guide_hysteresis.pdf"
if (!file.exists(fig)){
  library(ggplot2)
  x <- readRDS("./data/resp_summ5.rds")
  x$composition <- as.character(x$composition)
  x <- x[order(x$mean_mean_hyst),]
  x$composition <- factor(x$composition, x$composition, ordered = TRUE)

 p <- ggplot(x) +
  geom_point(aes(x = composition, y = mean_mean_hyst), shape = 95, size = 25) +
  xlab("Functional groups present") +
  ylab("Amount of hysteresis") +
  #ylim(0,2) +
  theme_light() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

 ggsave(
   filename = fig,
   plot = p,
   device = "pdf"
 )

}
knitr::include_graphics(fig)
```

@@

As discussed in the Introduction section, the package was largely motivated by the question of how biodiversity influences ecosystem responses to environmental change. Extensive results concerning this question are given in a separate publication @Limberger2022. Here we describe one of the results, which is that having biodiversity in a functional group can allow state changes to occur that otherwise would not have. I.e. biodiversity can qualitatively change the state of the ecosystem.

Biodiversity was added to the functional groups using the `new_strain_parameter()` function to create a parameter set with multiple species per functional group (albeit all with identical features) and then to add variability among the species by calling the `add_strain_var()` function. This function takes an already existing parameter set and adds the specified about of variation. The new parameter object is then used as before.

@@

**Rainer... please beautify the figure as required. It comes from the Limberger et al supplement, Figure 5.4**
**@Owen: is this OK like this? Used the graph from the supplement**

```{r uc3, echo = FALSE,  include=TRUE, fig.cap='The dynamics of the ecosystem when there are two species in each functional group, and some variation (diversity) in species parameters.', fig.align="center", out.width = 350 }
knitr::include_graphics("figures/uc3_supplement_5_4.png")
```

@@

*microxanox* allows the simulation, visualisation, and analysis of a model of a microbial ecosystem while allowing variation in the amount of diversity containing in each of the functional groups of organisms present. It has been used for the research described in another paper of ours that is one of the first investigations of the effects of diversity on ecosystem resilience @Limberger2022. In that paper, we show that diversity can have large and important effects of ecosystem responses, highlighting the need for models such as ours, within which one can easily manipulate the amount of biodiversity. The *microxanox* package has also been used to reproduce the results of the paper that inspired the package development @Bush2017.

The package greatly lowers the amount of work required in further investigations of the specific ecosystem modelled. There has, for example, been quite limited investigation of how biodiversity influences the short-term responses of the modelled ecosystem to environmental change. Likewise, the package could be used to power an investigation of the effects of biodiversity on the usefulness of early warning signals of abrupt ecosystem change [@Scheffer2009]. In addition this package could be used as a template for the implementation for developing models of other types of ecosystems and organism. By doing so, other models can profit from the overall framework used, and the reproducibility aspects as well as the flexibility implemented.
